---
title: "Untitled"
output: html_document
date: "2023-02-07"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(TSCAN)
library(tidyverse)
library(dbscan)
library(pscl)
library(matrixStats)
library(dtwclust)
library(dtw)
library(dplyr)
library(ts2net)
```

```{r}
dat = read.csv("exp_mat_tp.csv") #expression matrix
pt = read.csv("lpsorder.csv", row.names = 1) #pseudotime order
```

##Use lowess regression to smooth trajectories 

```{r}
dat_pt = dat[match(pt[,1], dat$X),]
dat_pt$pt = c(1:length(pt[,1]))

##Observe the overall trajectories plots
matrix_all = rep(NA, 14)
for (i in 3:16) {
  model = lowess(dat_pt$pt, dat_pt[,i])[2]
  matrix_all[i] = model
}
lowess_all = as.data.frame(matrix_all[3:16])
gene = colnames(dat[3:16])
colnames(lowess_all) = gene

col_set <- rainbow(ncol(lowess_all))
matplot(c(1:length(pt[,1])), lowess_all, type="l", col = col_set)

nn <- ncol(lowess_all)
layout(matrix(c(1,2),nrow=1), width=c(5,3)) 
par(mar=c(5,4,4,0)) #No margin on the right side
matplot(c(1:length(pt[,1])), lowess_all, type="l", col = col_set,
        xlab="Cell",ylab="Expression",main="Pseudo-time Trajectories for All Genes")
par(mar=c(5,0,4,2)) #No margin on the left side
plot(c(0,1),type="n", axes=F, xlab="", ylab="")
legend("center", gene,col=seq_len(nn),cex=0.8,fill=seq_len(nn))
```
## Remove markers with low vairations
```{r}
lowesss_mat_all = as.matrix(lowess_all)
colMaxs(lowesss_mat_all) - colMins(lowesss_mat_all)
colnames(lowesss_mat_all)

diff = data.frame(colnames(lowesss_mat_all),colMaxs(lowesss_mat_all) - colMins(lowesss_mat_all) )
colnames(diff) = c('gene', 'diff')
diff
```

## Selected markers with high variations
```{r}
sel_gene = diff %>% filter(diff > 200) 
sel_gene = sel_gene[,1]
```

```{r}
dat = dat[,c('X', 'timepoint', sel_gene)]
```

## Seperate expression matrix by 5 time points

```{r}
tp0 = dat %>% filter(timepoint == 'tp0')
tp1 = dat %>% filter(timepoint == 'tp1')
tp4 = dat %>% filter(timepoint == 'tp4')
tp7 = dat %>% filter(timepoint == 'tp7')
tp14 = dat %>% filter(timepoint == 'tp14')
```


```{r}
pt0 = pt[pt$x %in% tp0$X,]
pt1 = pt[pt$x %in% tp1$X,]
pt4 = pt[pt$x %in% tp4$X,]
pt7 = pt[pt$x %in% tp7$X,]
pt14 = pt[pt$x %in% tp14$X,]
```

```{r}
tp0_pt = tp0[match(pt0, tp0$X),]
tp0_pt$pt = c(1:length(pt0))

tp1_pt = tp1[match(pt1, tp1$X),]
tp1_pt$pt = c(1:length(pt1))

tp4_pt = tp4[match(pt4, tp4$X),]
tp4_pt$pt = c(1:length(pt4))

tp7_pt = tp7[match(pt7, tp7$X),]
tp7_pt$pt = c(1:length(pt7))

tp14_pt = tp14[match(pt14, tp14$X),]
tp14_pt$pt = c(1:length(pt14))
```

## Time point 0
```{r}
# Lowess
matrix0 = rep(NA, 5)
for (i in 3:dim(tp0_pt)[2]) {
  model = lowess(tp0_pt$pt, tp0_pt[,i])[2]
  matrix0[i] = model
}
lowess_tp0 = as.data.frame(matrix0[3:7])
colnames(lowess_tp0) = sel_gene

#Plots
nn <- ncol(lowess_tp0)
layout(matrix(c(1,2),nrow=1), width=c(4,2)) 
par(mar=c(5,4,4,0)) #No margin on the right side
matplot(c(1:length(pt0)), lowess_tp0, type="l",
        xlab="Cell",ylab="Expression",main="Trajectories for time point 1")
par(mar=c(5,0,4,2)) #No margin on the left side
plot(c(0,1),type="n", axes=F, xlab="", ylab="")
legend("center", sel_gene,col=seq_len(nn),cex=0.8,fill=seq_len(nn))
```


## Time point 1
```{r}
# Lowess
matrix1 = rep(NA, 5)
for (i in 3:7) {
  model = lowess(tp1_pt$pt, tp1_pt[,i])[2]
  matrix1[i] = model
}
lowess_tp1 = as.data.frame(matrix1[3:7])
colnames(lowess_tp1) = sel_gene

#Plots
nn <- ncol(lowess_tp1)
layout(matrix(c(1,2),nrow=1), width=c(4,2)) 
par(mar=c(5,4,4,0)) #No margin on the right side
matplot(c(1:length(pt1)), lowess_tp1, type="l",
        xlab="Cell",ylab="Expression",main="Trajectories for time point 2")
par(mar=c(5,0,4,2)) #No margin on the left side
plot(c(0,1),type="n", axes=F, xlab="", ylab="")
legend("center", sel_gene,col=seq_len(nn),cex=0.8,fill=seq_len(nn))
```

## Time point 4
```{r}
matrix4 = rep(NA, 7)
for (i in 3:7) {
  model = lowess(tp4_pt$pt, tp4_pt[,i])[2]
  matrix4[i] = model
}
lowess_tp4 = as.data.frame(matrix4[3:7])
colnames(lowess_tp4) = sel_gene


#Plots
nn <- ncol(lowess_tp4)
layout(matrix(c(1,2),nrow=1), width=c(4,2)) 
par(mar=c(5,4,4,0)) #No margin on the right side
matplot(c(1:length(pt4)), lowess_tp4, type="l",
        xlab="Cell",ylab="Expression",main="Trajectories for time point 3")
par(mar=c(5,0,4,2)) #No margin on the left side
plot(c(0,1),type="n", axes=F, xlab="", ylab="")
legend("center", sel_gene,col=seq_len(nn),cex=0.8,fill=seq_len(nn))
```

## Time point 7
```{r}
# Lowess
matrix7 = rep(NA, 7)
for (i in 3:7) {
  model = lowess(tp7_pt$pt, tp7_pt[,i])[2]
  matrix7[i] = model
}
lowess_tp7 = as.data.frame(matrix7[3:7])
colnames(lowess_tp7) = sel_gene

#Plots
nn <- ncol(lowess_tp7)
layout(matrix(c(1,2),nrow=1), width=c(4,2)) 
par(mar=c(5,4,4,0)) #No margin on the right side
matplot(c(1:length(pt7)), lowess_tp7, type="l",
        xlab="Cell",ylab="Expression",main="Trajectories for time point 4")
par(mar=c(5,0,4,2)) #No margin on the left side
plot(c(0,1),type="n", axes=F, xlab="", ylab="")
legend("center", sel_gene,col=seq_len(nn),cex=0.8,fill=seq_len(nn))
```

## Time point 14
```{r}
# Lowess
matrix14 = rep(NA, 7)
for (i in 3:7) {
  model = lowess(tp14_pt$pt, tp14_pt[,i])[2]
  matrix14[i] = model
}
lowess_tp14 = as.data.frame(matrix14[3:7])
colnames(lowess_tp14) = sel_gene


#Plots
nn <- ncol(lowess_tp14)
layout(matrix(c(1,2),nrow=1), width=c(4,2)) 
par(mar=c(5,4,4,0)) #No margin on the right side
matplot(c(1:length(pt14)), lowess_tp14, type="l",
        xlab="Cell",ylab="Expression",main="Trajectories for time point 5")
par(mar=c(5,0,4,2)) #No margin on the left side
plot(c(0,1),type="n", axes=F, xlab="", ylab="")
legend("center", sel_gene,col=seq_len(nn),cex=0.8,fill=seq_len(nn))
```

## DTW for Marker/Gene 
```{r}
#dtw distance normalization function
ndtw <- function(x, y = NULL, ...) {
    dtw::dtw(x, y, ..., distance.only = TRUE)$normalizedDistance
}

pr_DB$set_entry(
  FUN = ndtw,
  names = "ndtw",
  loop = TRUE,
  distance = TRUE
)
```

## Normalize Trajectories and dtw distances
```{r}
df_tra0 = lowess_tp0 %>% mutate_all(funs((. - min(., na.rm = T))/(max(., na.rm = T) - min(., na.rm = T)))) #normalization of trjectories
d0 = proxy::dist(t(df_tra0), method = "ndtw", upper = TRUE, diag = TRUE) #dtw distance normalization

df_tra1 = lowess_tp1 %>% mutate_all(funs((. - min(., na.rm = T))/(max(., na.rm = T) - min(., na.rm = T))))
d1 = proxy::dist(t(df_tra1), method = "ndtw", upper = TRUE, diag = TRUE)


df_tra4 = lowess_tp4 %>% mutate_all(funs((. - min(., na.rm = T))/(max(., na.rm = T) - min(., na.rm = T))))
d4 = proxy::dist(t(df_tra4), method = "ndtw", upper = TRUE, diag = TRUE)

df_tra7 = lowess_tp7 %>% mutate_all(funs((. - min(., na.rm = T))/(max(., na.rm = T) - min(., na.rm = T))))
d7 = proxy::dist(t(df_tra7), method = "ndtw", upper = TRUE, diag = TRUE)

df_tra14 = lowess_tp14 %>% mutate_all(funs((. - min(., na.rm = T))/(max(., na.rm = T) - min(., na.rm = T))))
d14 = proxy::dist(t(df_tra14), method = "ndtw", upper = TRUE, diag = TRUE)
```

## Calculate matrix as input for GenLouvain
```{r}
mat_d0 = 1 - as.matrix(d0) 
diag(mat_d0) = 0
mat_d1 = 1 - as.matrix(d1)
diag(mat_d1) = 0
mat_d4 = 1 - as.matrix(d4)
diag(mat_d4) = 0
mat_d7 = 1 - as.matrix(d7)
diag(mat_d7) = 0
mat_d14 = 1 - as.matrix(d14)
diag(mat_d14) = 0

#write.csv(mat_d14, "mat_d14.csv", row.names = F)
```

## Distance matrix for 1,000 cells
```{r}
interval0 = floor(seq.int(from=1,to=nrow(lowess_tp0),length.out=1000)) #subsetting data
cell_tr0_300 = lowess_tp0[interval0,]  %>% 
  mutate_all(funs((. - min(., na.rm = T))/(max(., na.rm = T) - min(., na.rm = T)))) #normalization 
cell_d0 = dist(cell_tr0_300, method = "euclidean") #distance matrix

interval1 = floor(seq.int(from=1,to=nrow(lowess_tp1),length.out=1000))
cell_tr1_300 = lowess_tp1[interval1,] %>% 
  mutate_all(funs((. - min(., na.rm = T))/(max(., na.rm = T) - min(., na.rm = T))))
cell_d1 = dist(cell_tr1_300, method = "euclidean")

interval4 = floor(seq.int(from=1,to=nrow(lowess_tp4),length.out=1000))
cell_tr4_300 = lowess_tp4[interval4,] %>% 
  mutate_all(funs((. - min(., na.rm = T))/(max(., na.rm = T) - min(., na.rm = T))))
cell_d4 = dist(cell_tr4_300, method = "euclidean")

interval7 = floor(seq.int(from=1,to=nrow(lowess_tp7),length.out=1000))
cell_tr7_300 = lowess_tp7[interval7,] %>% 
  mutate_all(funs((. - min(., na.rm = T))/(max(., na.rm = T) - min(., na.rm = T))))
cell_d7 = dist(cell_tr7_300, method = "euclidean")

interval14 = floor(seq.int(from=1,to=nrow(lowess_tp14),length.out=1000))
cell_tr14_300 = lowess_tp14[interval14,] %>% 
  mutate_all(funs((. - min(., na.rm = T))/(max(., na.rm = T) - min(., na.rm = T))))
cell_d14 = dist(cell_tr14_300, method = "euclidean")

```

## Euclidean distance normalization
```{r}
n_cell_d0 = dist_matrix_normalize(as.matrix(cell_d0), to = c(0, 1))
n_cell_d1 = dist_matrix_normalize(as.matrix(cell_d1), to = c(0, 1))
n_cell_d4 = dist_matrix_normalize(as.matrix(cell_d4), to = c(0, 1))
n_cell_d7 = dist_matrix_normalize(as.matrix(cell_d7), to = c(0, 1))
n_cell_d14 = dist_matrix_normalize(as.matrix(cell_d14), to = c(0, 1))

cell_mat_d0 = 1 - n_cell_d0
diag(cell_mat_d0) = 0
cell_mat_d1 = 1 - n_cell_d1
diag(cell_mat_d1) = 0
cell_mat_d4 = 1 - n_cell_d4
diag(cell_mat_d4) = 0
cell_mat_d7 = 1 - n_cell_d7
diag(cell_mat_d7) = 0
cell_mat_d14 = 1 - n_cell_d14
diag(cell_mat_d14) = 0

#write.csv(cell_mat_d0, "cell_mat_d0.csv", row.names = F)
#write.csv(cell_mat_d1, "cell_mat_d1.csv", row.names = F)
#write.csv(cell_mat_d4, "cell_mat_d4.csv", row.names = F)
#write.csv(cell_mat_d7, "cell_mat_d7.csv", row.names = F)
#write.csv(cell_mat_d14, "cell_mat_d14.csv", row.names = F)
```




```{r}
matplot(c(1:length(pt0)), lowess_tp0, type="l",
        xlab="Cell",ylab="Expression",main="Trajectories for time point 0")
matplot(c(1:length(pt1)), lowess_tp1, type="l",
        xlab="Cell",ylab="Expression",main="Trajectories for time point 1")
matplot(c(1:length(pt4)), lowess_tp4, type="l",
        xlab="Cell",ylab="Expression",main="Trajectories for time point 4")
matplot(c(1:length(pt7)), lowess_tp7, type="l",
        xlab="Cell",ylab="Expression",main="Trajectories for time point 7")
matplot(c(1:length(pt14)), lowess_tp14, type="l",
        xlab="Cell",ylab="Expression",main="Trajectories for time point 14")
```










